---
title: "Introduction to causal mediation"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to causal mediation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette introduces `mediator`and provides an overview for mediation analysis under the counterfactual framework. 

***

## Terminology -- currently taken from Valeri and VanderWeele

**Controlled Direct Effect (CDE)** expresses how much the outcome would change on average if the mediator were controlled at level *m* uniformly in the population but the treatment were changed from level *a*=0 to level *a*=1

**Natural Direct Effect (NDE)** expresses how much the outcome would change if the exposure were set at level *a*=1 versus level *a*=0 but for each individual the mediatior were kept at the level it woud have taken in the absense of the exposure. 

**Natural Indirect Effect (NIE)** expresses how much the outcome would change on average if the exposure were controlled at level *a*=1, but the mediator were changed from the level if would take if *a*=0 to the level it would take if *a*=1. 

**Total Effect** how much the outcome would change overall for a change in the exposure from level *a*=0 to level *a*=1.  

**Proportion Mediated** the extent to which the total effect of the exposure on the outcome operates through the mediator

## Comparison to other tools 

`mediator` is related to the SAS/SPSS macros developed by Valeri and VanderWeele, however minor differences exist between the programs which sometimes results in variation in estimates and confidence intervals. 

When the macros calculate the CI7 using the delta method, they use the hard coded values 1.96 and -1.96 while `mediator` uses `c(-1,1)*qnorm(.975)`. Languages also differ with respect to rounding. 

The SAS macro takes roughly 7.05 seconds to run, calculating the CI using the delta method, and roughly 276.55 seconds using bootstrapping, while `mediator` takes 0.01 seconds for the delta method and 0.27 seconds for bootstrapping (using a dataset with 250 observations and 100 replicates). 

When using bootstrapping the SAS macro bootstraps both the effect estimates and CI, while the R function only uses the bootstrap for obtaining the CI. Users should use the bootstrap function with caution, especially when using small datasets. We recommend using a minimum 10000 replicates. Using more replicates will take more time, however it will provide more precise estimates. 

```{r setup}
#library(mediator)
```


## Equations

For a continuous outcome and mediator: 

Models for Y (outcome) and M (mediator), correctly specified

<center>

![](https://latex.codecogs.com/gif.latex?E%5BY%7Ca%2Cm%2Cc%5D%20%3D%20%5Ctheta_%7B0%7D%20&plus;%20%5Ctheta_%7B1%7Da%20&plus;%20%5Ctheta_%7B2%7Dm%20&plus;%20%5Ctheta_%7B3%7Dam%20&plus;%20%5Ctheta%27_%7B4%7Dc)

![](https://latex.codecogs.com/gif.latex?E%5BM%7Ca%2Cc%5D%20%3D%20%5Cbeta_%7B0%7D%20&plus;%20%5Cbeta_%7B1%7Da%20&plus;%20%5Cbeta%27_%7B2%7Dc)

</center>

Average controlled direct effect, natural direct effect and natural indirect effect 

<center>

![](https://latex.codecogs.com/gif.latex?E%5BY_%7Bam%7D%20-%20Y_%7Ba*m%7D%7Cc%5D%20%3D%20%28%5Ctheta_%7B1%7D%20&plus;%20%5Ctheta_%7B3%7Dm%29%28a-a%5E%7B*%7D%29)

![](https://latex.codecogs.com/gif.latex?E%5BY_%7BaM_%7Ba%5E%7B*%7D%7D%7D%20-%20Y_%7Ba%5E%7B*%7DM_%7Ba%5E%7B*%7D%7D%7D%7Cc%5D%3D%5C%7B%5Ctheta_%7B1%7D%20&plus;%20%5Ctheta_%7B3%7D%28%5Cbeta_%7B0%7D%20&plus;%20%5Cbeta_%7B1%7Da%5E%7B*%7D&plus;%20%5Cbeta%27_%7B2%7Dc%29%5C%7D%28a-a%5E%7B*%7D%29)

![](https://latex.codecogs.com/gif.latex?E%5BY_%7BaM_%7Ba%7D%7D-Y_%7BaM_%7Ba%5E%7B*%7D%7D%7D%7Cc%5D%20%3D%20%28%5Ctheta_%7B2%7D%5Cbeta_%7B1%7D%20&plus;%20%5Ctheta_%7B3%7D%5Cbeta_%7B1%7Da%29%28a-a%5E%7B*%7D%29)

</center>

Binary outcome and continuous mediator

<center>

![](https://latex.codecogs.com/gif.latex?logit%5C%7BP%28Y%3D1%7Ca%2Cm%2Cc%29%5C%7D%3D%5Ctheta_%7B0%7D&plus;%5Ctheta_%7B1%7Da%20&plus;%20%5Ctheta_%7B2%7Dm%20&plus;%20%5Ctheta_%7B3%7Dam%20&plus;%20%5Ctheta%27_%7B4%7Dc)

![](https://latex.codecogs.com/gif.latex?E%5BM%7Ca%2Cc%5D%3D%20%5Cbeta_%7B0%7D%20&plus;%20%5Cbeta_%7B1%7Da%20&plus;%20%5Cbeta%27_%7B2%7Dc)

![](https://latex.codecogs.com/gif.latex?OR%5E%7BCDE%7D%28m%29%20%3D%20exp%5C%7B%28%5Ctheta_%7B1%7D&plus;%5Ctheta_%7B3%7Dm%29%28a-a%5E%7B*%7D%29%5C%7D)

![](https://latex.codecogs.com/gif.latex?OR%5E%7BNDE%7D%20%3D%20exp%5C%7B%28%5Ctheta_%7B1%7D&plus;%5Ctheta_%7B3%7D%5Cbeta_%7B0%7D&plus;%5Ctheta_%7B3%7D%5Cbeta_%7B1%7Da%5E%7B*%7D&plus;%5Ctheta_%7B3%7D%5Cbeta%27_%7B2%7DC&plus;%5Ctheta_%7B3%7D%5Ctheta_%7B2%7D%5Csigma%5E%7B2%7D%29%28a-a%5E%7B2%7D%29%5C%7D)

![](https://latex.codecogs.com/gif.latex?OR%5E%7BNIE%7D%20%3D%20exp%5C%7B%28%5Ctheta_%7B2%7D%5Cbeta_%7B1%7D%20&plus;%20%5Ctheta_%7B3%7D%5Cbeta_%7B1%7Da%29%28a-a%5E%7B*%7D%29%5C%7D)

</center>

Continuous outcome and binary mediator

<center>

![](https://latex.codecogs.com/gif.latex?E%5BY%7Ca%2Cm%2Cc%5D%20%3D%20%5Ctheta_%7B0%7D%20&plus;%20%5Ctheta_%7B1%7Da%20&plus;%20%5Ctheta_%7B2%7Dm%20&plus;%20%5Ctheta_%7B3%7Dam%20&plus;%20%5Ctheta%27_%7B4%7Dc)

![](https://latex.codecogs.com/gif.latex?logit%5C%7BP%28M%3D1%7Ca%2Cc%29%5C%7D%3D%5Cbeta_%7B0%7D&plus;%5Cbeta_%7B1%7Da&plus;%5Cbeta%27_%7B2%7Dc)

![](https://latex.codecogs.com/gif.latex?E%5BY_%7Bam%7D-Y_%7Ba%5E%7B*%7Dm%7D%7Cc%5D%20%3D%20%28%5Ctheta_%7B1%7D&plus;%5Ctheta_%7B3%7Dm%29%28a-a%5E%7B*%7D%29)

![](https://latex.codecogs.com/gif.latex?E%5BY_%7BaM_%7Ba%5E%7B*%7D%7D%7D-Y_%7Ba%5E%7B*%7DM_%7Ba%5E%7B*%7D%7D%7D%7Cc%5D%3D%5C%7B%5Ctheta_%7B1%7D%28a-a%5E%7B*%7D%29%5C%7D%20&plus;%20%5C%7B%5Ctheta_%7B3%7D%28a-a%5E%7B*%7D%29%5C%7D%5Cdfrac%7Bexp%5B%5Cbeta_%7B0%7D&plus;%5Cbeta_%7B1%7Da%5E%7B*%7D&plus;%5Cbeta%27_%7B2%7Dc%5D%7D%7B1&plus;exp%5B%5Cbeta_%7B0%7D&plus;%5Cbeta_%7B1%7Da%5E%7B*%7D&plus;%5Cbeta%27_%7B2%7Dc%5D%7D)

![](https://latex.codecogs.com/gif.latex?E%5BY_%7BaM_%7Ba%7D%7D-Y_%7BaM_%7Ba%5E%7B*%7D%7D%7D%7Cc%5D%3D%28%5Ctheta_%7B2%7D&plus;%5Ctheta_%7B3%7Da%29%20%5C%7B%5Cdfrac%7Bexp%5B%5Cbeta_%7B0%7D&plus;%5Cbeta_%7B1%7Da&plus;%5Cbeta%27_%7B2%7Dc%5D%7D%7B1&plus;exp%5B%5Cbeta_%7B0%7D&plus;%5Cbeta_%7B1%7Da&plus;%5Cbeta%27_%7B2%7Dc%5D%7D-%5Cdfrac%7Bexp%5B%5Cbeta_%7B0%7D&plus;%5Cbeta_%7B1%7Da%5E%7B*%7D&plus;%5Cbeta%27_%7B2%7Dc%5D%7D%7B1&plus;exp%5B%5Cbeta_%7B0%7D&plus;%5Cbeta_%7B1%7Da%5E%7B*%7D&plus;%5Cbeta%27_%7B2%7Dc%5D%7D%5C%7D)

</center>

Binary outcome and mediator

<center>

![](https://latex.codecogs.com/gif.latex?logit%5C%7BP%28Y%3D1%7Ca%2Cm%2Cc%29%5C%7D%20%3D%20%5Ctheta_%7B0%7D%20&plus;%20%5Ctheta_%7B1%7Da&plus;%20%5Ctheta_%7B2%7Dm%20&plus;%20%5Ctheta_%7B3%7Dam%20&plus;%20%5Ctheta%27_%7B4%7Dc)

![](https://latex.codecogs.com/gif.latex?logit%5C%7BP%28M%3D1%7Ca%2Cc%29%5C%7D%20%3D%20%5Cbeta_%7B0%7D%20&plus;%20%5Cbeta_%7B1%7Da%20&plus;%20%5Cbeta%27_%7B2%7Dc)

![](https://latex.codecogs.com/gif.latex?OR%5E%7BCDE%7D%28m%29%3Dexp%5B%28%5Ctheta_1%20&plus;%20%5Ctheta_3m%29%28a-a%5E*%29%5D)

![](https://latex.codecogs.com/gif.latex?OR%5E%7BNDE%7D%20%3D%20%5Cdfrac%7Bexp%28%5Ctheta_1a%29%5C%7B1&plus;exp%28%5Ctheta_2&plus;%5Ctheta_3a&plus;%5Cbeta_0&plus;%5Cbeta_1a%5E*&plus;%5Cbeta%27_2c%29%5C%7D%7D%7Bexp%28%5Ctheta_1a%5E*%29%5C%7B1&plus;exp%28%5Ctheta_2&plus;%5Ctheta_3a%5E*&plus;%5Cbeta_0&plus;%5Cbeta_1a%5E*&plus;%5Cbeta%27_2c%29%5C%7D%7D)

![](https://latex.codecogs.com/gif.latex?OR%5E%7BNIE%7D%20%3D%20%5Cdfrac%7B%5C%7B1&plus;exp%28%5Cbeta_0&plus;%5Cbeta_1a%5E*&plus;%5Cbeta%27_2c%29%5C%7D%5C%7B1&plus;exp%28%5Ctheta_2&plus;%5Ctheta_3a&plus;%5Cbeta_0&plus;%5Cbeta_1a&plus;%5Cbeta%27_2c%29%5C%7D%7D%7B%5C%7B1&plus;exp%28%5Cbeta_0&plus;%5Cbeta_1a&plus;%5Cbeta%27_2c%29%5C%7D%5C%7B1&plus;exp%28%5Ctheta_2&plus;%5Ctheta_3a&plus;%5Cbeta_0&plus;%5Cbeta_1a%5E*&plus;%5Cbeta%27_2c%29%5C%7D%7D)

</center>
